Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-04-20 demystify.tex
updates

However, if your goal is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. A non-CS faculty can serve in the PhD dissertation committee (common) or \emph{co-advise} (somewhat rare), but your main PhD adviser will likely be a tenure-line faculty in CS (\autoref{sec:faculty-types}). 
For example, a prof. in Stats or Math dept might be able to serve as a co-adviser, but not as a sole adviser of a student in a CS PhD program.  Similarly, a prof. in CS cannot be the sole adviser and graduate a PhD student in a Math or Stats.
If in doubt, check with the CS department for their requirements.

@@

A related reason is that you did not show enough interest in the program.
For example, your SOP did not mention any faculty or research area at the program, you did not respond to professors interested in interviewing you (which might be considered unprofessional and burn bridges), or during the interview you show little knowledge, interest, or enthusiasm (because they are your safety schools) in the program or profs.

\paragraph{Red flags in Your Background} Plagiarism, cheating, or other academic dishonesty are \red{red} flags. Another one is if you have a history of jumping from one program to another (e.g., you have been in multiple different programs in a short time). How would the reviewers know?  They might have contacts in other programs and heard about you, your LoRs might mention it.

@@

%\section{How exactly do you evaluate an application?}\label{sec:ievaluate}

>>>>

However, if your goal is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. A non-CS faculty can serve in the PhD dissertation committee (common) or \emph{co-advise} (somewhat rare), but your main PhD adviser will likely be a tenure-line faculty in CS (\autoref{sec:faculty-types}). 
For example, a prof. in Stats or Math might be able to serve as a co-adviser, but not as a sole adviser of a student in a CS PhD program.  Similarly, a prof. in CS cannot be the sole adviser and graduate a PhD student in Math or Stats.
If in doubt, check with the CS department for their requirements.

@@

A related reason is that you did not show enough interest in the program.
For example, your SOP did not mention any faculty or research area at the program, you did not respond to professors interested in interviewing you (which might be considered unprofessional and burn bridges), or during the interview you showed little knowledge, interest, or enthusiasm (because they are your safety schools) in the program or profs.

\paragraph{Red flags} Various types of red flags can cause concern to the adcom (e.g., plagiarism, cheating, or other academic dishonesty). Another one is that you have a history of jumping from one program to another (e.g., you have been in multiple different programs in a short time). How would the reviewers know?  They might have contacts in other programs and heard about you, your LoRs might mention it.

@@

% \section{How long do you evaluate an application?}\label{sec:ievaluate}

% On average, I spend about 15 minutes reviewing each application (less for clear rejections and more for potential acceptances). In earlier years I spent more time but now  I’ve become more efficient in evaluating..

% The review system compiles a single PDF per applicant, including a summary page, transcripts, test scores, LoRs, a CV, SOP, and any writing samples. I review these documents in order, taking notes to support my evaluations.

% I start with the summary, checking for low GPAs or test scores below the university minumum. I then skim the transcripts for low grades in relevant courses, noting issues like "- many low grades in main courses and - unknown international school with good GPA".

% I pay close attention to strong LoRs and skim weaker ones, noting either strong endorsements from well-known professors or generic content. For the CV, I look for publications, research experiences, and notable achievements, noting things like "+ published papers in top venues" or "+ gold medal in an international competition."

% The SOP is read thoroughly unless it's poorly written; a brief skim suffices in those cases. I note whether the SOP is engaging and well-tailored to our program or if it’s generic and uninspired.

% Finally, I make a decision—usually either a direct rejection or an offer of admission with full funding, mentioning potential funding sources. Exceptionally strong candidates are recommended for the University Presidential fellowship. The admissions committee chair reviews all evaluations, typically following the recommendations, but discrepancies are resolved through discussion among reviewers to reach a consensus.
